MR-PFU: A suite of R packages for energy conversion chain analysis

### 2023-07-24 paper.Rmd
More updates to the paper.

@@

Energy flows through societal energy conversion chains (ECCs)
to enable economic activity and facilitate human flourishing.
To understand economic growth and human well-being, 
the field of energy analysis evaluates ECCs
from the primary stage 
(resources extracted from the environment, 
such as coal, oil, natural gas) 
to the final stage
(energy purchased by consumers, 
such as refined petroleum and electricity),
to the useful stage
(energy desired by the end user, 
such as heat, motion, and light),
and sometimes to energy services
(such as thermal comfort, transport, and illumination).
Societal exergy analysis (SEA), 
an extension of energy analysis,
quantifies ECCs in exergy terms^[Exergy is 
the mechanical work potential of energy.].

@@

Historically, societal exergy analysts have
analyzed the ECCs of individual countries
using a large number of linked spreadsheets,
starting with primary- and final-stage data 
from the [IEA](https://www.iea.org/data-and-statistics/data-product/world-energy-balances)'s
world energy balances
[@Ayres:2003ec; @Serrenho:2014aa; @Brockway:2014aa; @Brockway:2015aa]. 
Data were in varying and inconsistent formats.
The authors of this paper and others in the field
wanted to expand SEA to cover all countries, but
the spreadsheet approach to SEA was
deemed not scalable.
We created a suite of `R` packages to create a database^[Strictly speaking, we create data frames of matrices, 
not a structured database.]
of PFU data (the PFU Database)
for all countries in the world from 1960--2020.
The packages enable, for the first time, scalable SEA.
This paper describes the design of those packages and
demonstrates briefly their use.

@@

Further development followed from selection of the PSUT framework.
First, the matrices of the PSUT framework carry the challenge that 
different countries and years 
have varying energy carriers (products) and
varying energy conversion machines (industries),
meaning that PSUT matrices for different countries and years 
have differing sizes and differing row and column names.
To get around this challenge, we created the
[matsbyname](https://github.com/MatthewHeun/matsbyname/) [@Heun-matsbyname:2023]
package which enables matrix mathematics 
that respects matrix row and column names, 
inserting rows or columns of `0`s when needed.
Second,
we knew it would be convenient to perform _matrix_ mathematics 
as easily as _scalar_ mathematics
in `R` data frames
using the [tidyverse](https://www.tidyverse.org) syntax [@Wickham:2019]. 
We developed the [matsindf](https://matthewheun.github.io/matsindf/) package [@Heun-matsindf:2023]
to enable this functionality.
Finally, manipulating row and column names proved to be a challenge, 
especially for matrices in PSUT data frames, so 
we developed the [RCLabels](https://matthewheun.github.io/RCLabels/) package [@Heun-RCLabels:2023]
for that purpose.
The table below summarizes these packages, 
all of which are generally useful and available on 
[CRAN](https://cran.r-project.org).

| Package | Function |
|:--------|:---------|
| [RCLabels](https://github.com/MatthewHeun/RCLabels/)      | Manipulates row and column names in [matsindf](https://github.com/MatthewHeun/matsindf/) data frames |
| [matsbyname](https://github.com/MatthewHeun/matsbyname/)  | Performs matrix mathematics that respects row and column names |
| [matsindf](https://github.com/MatthewHeun/matsindf/)      | Stores matrices in cells of a data frame, thereby enabling [tidyverse](https://www.tidyverse.org) syntax |

Several calculation steps are required to create the PFU database. 
Each step is assisted by an `R` package that we created.
First, the IEA's primary- and final-stage WEEB data must be converted
to the PSUT format, 
a task completed by the [IEATools](https://github.com/MatthewHeun/IEATools/) package [@Heun-IEATools:2023]. 
Second, human and animal muscle work must be calculated from 
labor and FAO data following the methodology of @Steenwyk:2022ww
using the [MWTools](https://github.com/EnergyEconomyDecoupling/MWTools) package [@Marshall:2023ab]. 
Third, the IEA's primary- and final-stage ECC data 
are extended to the useful stage by
(a) allocating final stage energy to end-use machines and
(b) multiplying allocated final energy by the 
final-to-useful efficiency of each machine.
This task is accomplished by the
[Recca](https://github.com/MatthewHeun/Recca) [@Heun-Recca:2023]
and 
[PFUDatabase](https://github.com/energyeconomydecoupling/PFUDatabase/) [@Heun-PFUDatabase:2023]
packages.
Finally, ECCs must be converted from energy terms to exergy terms. 
This step is assisted by the [Recca](https://github.com/MatthewHeun/Recca) package.
The steps to create the PSUT matrices for each country and each year
are accomplished by a
[targets](https://docs.ropensci.org/targets/) [@Landau:2021aa]
computation pipeline
available in the [PFUDatabase](https://github.com/energyeconomydecoupling/PFUDatabase/)
package.

A second [targets](https://docs.ropensci.org/targets/) pipeline 
in the [PFUAggDatabase](https://github.com/EnergyEconomyDecoupling/PFUAggDatabase) package [@Heun-PFUAggDatabase:2023]
aggregates ECCs
by region 
(continents and world),
by product category 
(e.g., Coal and coal products, Oil and oil products, Low-temperature heat, etc.), 
by end use categories
(e.g., Residential, Transport, etc.), and
to primary, final, and useful stages.
Furthermore, 
[PFUAggDatabase](https://github.com/EnergyEconomyDecoupling/PFUAggDatabase)
calculates aggregated efficiencies.
The [Recca](https://github.com/MatthewHeun/Recca) package is used extensively by the
[PFUAggDatabase](https://github.com/EnergyEconomyDecoupling/PFUAggDatabase) 
pipeline.

@@

| Package | Function |
|:--------|:---------|
| [IEATools](https://github.com/MatthewHeun/IEATools/) | Converts IEA data to [matsindf](https://github.com/MatthewHeun/matsindf/) format |
| [MWTools](https://github.com/energyeconomydecoupling/MWTools) | Converts ILO and FAO data to human and animal muscle work information in [matsindf](https://github.com/MatthewHeun/matsindf/) format |
| [Recca](https://matthewheun.github.io/Recca/)       | Performs `R` energy conversion chain analysis |
| [PFUDatabase](https://energyeconomydecoupling.github.io/PFUDatabase/) | A [targets](https://docs.ropensci.org/targets/) pipeline to create a data frame of PSUT matrices |
| [PFUAggDatabase](https://energyeconomydecoupling.github.io/PFUAggDatabase/) | A [targets](https://docs.ropensci.org/targets/) pipeline to aggregate PSUT matrices |

Reference for data repository [@Marshall:2023aa].

`Gala` is an Astropy-affiliated Python package for galactic dynamics. Python
enables wrapping low-level languages (e.g., C) for speed without losing
flexibility or ease-of-use in the user-interface. The API for `Gala` was
designed to provide a class-based and user-friendly interface to fast (C or
Cython-optimized) implementations of common operations such as gravitational
potential and force evaluation, orbit integration, dynamical transformations,
and chaos indicators for nonlinear dynamics. `Gala` also relies heavily on and
interfaces well with the implementations of physical units and astronomical
coordinate systems in the `Astropy` package [@astropy] (`astropy.units` and
`astropy.coordinates`).

`Gala` was designed to be used by both astronomical researchers and by
students in courses on gravitational dynamics or astronomy. It has already been
used in a number of scientific publications [@Pearson:2017] and has also been
used in graduate courses on Galactic dynamics to, e.g., provide interactive
visualizations of textbook material [@Binney:2008]. The combination of speed,
design, and support for Astropy functionality in `Gala` will enable exciting
scientific explorations of forthcoming data releases from the *Gaia* mission
[@gaia] by students and experts alike.

@@

>>>>

<!-- Define some macros to use later -->
[link-cran]:           https://cran.r-project.org
[link-ieatools]:       https://github.com/MatthewHeun/IEATools
[link-ieaweeb]:        https://www.iea.org/data-and-statistics/data-product/world-energy-balances
[link-matsbyname]:     https://github.com/MatthewHeun/matsbyname
[link-matsindf]:       https://github.com/MatthewHeun/matsindf
[link-mwtools]:        https://github.com/EnergyEconomyDecoupling/MWTools
[link-pfudatabase]:    https://github.com/energyeconomydecoupling/PFUDatabase
[link-pfuaggdatabase]: https://github.com/EnergyEconomyDecoupling/PFUAggDatabase
[link-rclabels]:       https://github.com/MatthewHeun/RCLabels
[link-recca]:          https://github.com/MatthewHeun/Recca
[link-targets]:        https://docs.ropensci.org/targets
[link-tidyverse]:      https://www.tidyverse.org

@@

Energy flowing through societal energy conversion chains (ECCs)
enables economic activity and facilitates human flourishing.
To understand economic growth and human well-being, 
the field of societal energy analysis evaluates ECCs
from the primary stage 
(resources extracted from the environment, 
such as coal, oil, and natural gas) 
to the final stage
(energy purchased by consumers, 
such as refined petroleum and electricity),
to the useful stage
(energy desired by the end user, 
such as heat, motion, and light),
and sometimes to energy services
(such as thermal comfort, transport, and illumination).
Societal exergy analysis (SEA), 
an extension of energy analysis,
quantifies ECCs in exergy terms^[Exergy is 
the mechanical work potential of energy.].

We created a suite of `R` packages 
(some of which are generally useful and available through 
[CRAN][link-cran],
others of which are focused on SEA and available on GitHub)
to assist SEA practitioners as they analyze 
energy movement through society. 
These new packages enable analysis of every country in the world
across timespans of decades or longer.
The new packages were used to create the "PFU Database," 
a new resource for the SEA community. 
The packages can be used for specific SEA analyses, too.

@@

Historically, societal exergy analysts have
analyzed the ECCs of individual countries
using a large number of linked spreadsheets,
often starting with primary- and final-stage data 
from the [IEA][link-ieaweeb]'s
world energy balances
[@Ayres:2003ec; @Serrenho:2014aa; @Brockway:2014aa; @Brockway:2015aa]. 
Data were in varying and inconsistent formats.
The authors of this paper and others in the field
wanted to expand SEA to cover all countries, but
the spreadsheet approach to SEA was
deemed not scalable.
We created a suite of `R` packages to create a database^[Strictly speaking, we create data frames of matrices, 
not a structured database.]
of PFU data (the PFU Database)
for all countries in the world from 1960--2020.
The packages enable, for the first time, scalable SEA.
This paper describes the design of those packages and
demonstrates briefly their use.

@@

Further development followed from selection of the PSUT framework.
First, the matrices of the PSUT framework carry the challenge that 
different countries and years 
have varying energy carriers (products) and
varying energy conversion machines (industries),
meaning that PSUT matrices for different countries and years 
have differing sizes and differing row and column names.
To get around this challenge, we created the
[matsbyname][link-matsbyname] [@Heun-matsbyname:2023]
package which enables matrix mathematics 
that respects matrix row and column names, 
inserting rows or columns of `0`s when needed.
Second,
we knew it would be convenient to perform _matrix_ mathematics 
as easily as _scalar_ mathematics
in `R` data frames
using the [tidyverse][link-tidyverse] syntax [@Wickham:2019]. 
We developed the [matsindf][link-matsindf] package [@Heun-matsindf:2023]
to enable this functionality.
Finally, manipulating row and column names proved to be a challenge, 
especially for matrices in PSUT data frames, so 
we developed the [RCLabels][link-rclabels] package [@Heun-RCLabels:2023]
for that purpose.
The table below summarizes these packages, 
all of which are generally useful and available on 
[CRAN][link-cran].

| Package | Function |
|:--------|:---------|
| [RCLabels][link-rclabels]     | Manipulates row and column names in [matsindf][link-matsindf] data frames |
| [matsbyname][link-matsbyname] | Performs matrix mathematics that respects row and column names |
| [matsindf][link-matsindf]     | Stores matrices in cells of a data frame, thereby enabling [tidyverse][link-tidyverse] syntax |

Several calculation steps are required to create the PFU database. 
Each step is assisted by an `R` package that we created.
First, the IEA's primary- and final-stage WEEB data must be converted
to the PSUT format, 
a task completed by the [IEATools][link-ieatools] package [@Heun-IEATools:2023]. 
Second, human and animal muscle work must be calculated from 
labor and FAO data following the methodology of @Steenwyk:2022ww
using the [MWTools][link-mwtools] package [@Marshall:2023ab]. 
Third, the IEA's primary- and final-stage ECC data 
are extended to the useful stage by
(a) allocating final stage energy to end-use machines and
(b) multiplying allocated final energy by the 
final-to-useful efficiency of each machine.
This task is accomplished by the
[Recca][link-recca] [@Heun-Recca:2023]
and 
[PFUDatabase](https://github.com/energyeconomydecoupling/PFUDatabase/) [@Heun-PFUDatabase:2023]
packages.
The pipeline is modular, such that data for an country can be added at a later date
to improve the database. 
Finally, ECCs must be converted from energy terms to exergy terms. 
This step is assisted by the [Recca][link-recca] package.
The steps to create the PSUT matrices for each country and each year
are accomplished by a
[targets][link-targets] [@Landau:2021aa]
computation pipeline
available in the [PFUDatabase][link-pfudatabase]
package.
A unique feature of the
[PFUDatabase][link-pfudatabase]
pipeline is an exemplar system that allows
analyses to proceed when allocation or efficiency data for a country are unavailable.

A second [targets][link-targets] pipeline 
in the [PFUAggDatabase][link-pfuaggdatabase] package [@Heun-PFUAggDatabase:2023]
aggregates ECCs
by region 
(continents and world),
by product category 
(e.g., Coal and coal products, 
Low-, Medium-, and High-temperature heat, etc.), 
by end use sectors
(e.g., Residential, Transport, etc.), and
to primary, final, and useful (PFU) stages.
Furthermore, 
[PFUAggDatabase][link-pfuaggdatabase]
calculates aggregated efficiencies across 
the primary, final, and useful stages of all ECCs in the database.
The [Recca][link-recca] package is used extensively by the
[PFUAggDatabase][link-pfuaggdatabase]
pipeline.

@@

| Package | Function |
|:--------|:---------|
| [IEATools][link-idatools]             | Converts IEA data to [matsindf][link-matsindf] format |
| [MWTools][link-mwtools]               | Converts ILO and FAO data to human and animal muscle work information in [matsindf][link-matsindf] format |
| [Recca][link-recca]                   | Performs `R` energy conversion chain analysis |
| [PFUDatabase][link-pfudatabase]       | A [targets][link-targets] pipeline to create a data frame of PSUT matrices |
| [PFUAggDatabase][link-pfuaggdatabase] | A [targets][link-targets] pipeline to aggregate PSUT matrices |

Input data for the PFU Database can be found in @Marshall:2023aa.
Access to the PFU Database can be obtained by correspondence 
with author [PEB](mailto:P.E.Brockway@leeds.ac.uk) data^[Access to the PFU Database
is restricted to those who have access to the [IEA WEEB data][link-ieaweeb].].

@@

The new `R` packages

enables
